---
title: "Beyond Filter and Select: Data Dictionary"
author: "Naomi Loughiln"
date: "2025-05-14"
output: html_document
---

# Beyond Filter and Select: Data Dictionaries 

Here we will create a data dictionary for the "Palmer Penguins" data set (a smaller dataset than the JumpStart dataset). The same concepts can be applied to the Jumpstart dataset.  

```{r setup, include=FALSE}
# install.packages("palmerpenguins")
``` 

# Creating a data dictionary for the penguins data set 

We would want to include  variable names, brief description, data type, labels, number of records or number of nas, ranges 

## Data
```{r}
penguins_data <- palmerpenguins::penguins
```

## Variable Names
```{r}
varNames <- names(penguins_data) 
```

## Data Types

Class reports on data structure (numeric, factor, etc.), typeof reports on the individual values (type or storage mode) of each object (e.g. double, integer, character).

```{r}
dataClass <- sapply(penguins_data, class) # apply outputs a list, whereas sapply outputs a simplified verion of a vector
view(as.data.frame(dataClass)) # verify our output 

dataType <- sapply(penguins_data, typeof)
view(as.data.frame(dataType))

```

## Missing Data
```{r}
observationsWithValues <- sapply(penguins_data, function(x) sum(!is.na(x)))
```

When using a nested function like is.na() within sum(), it can get confusing. We need to specify what we want apply is.na() on, so we make a temporary function using an 'x' as the variable to clarify the input to the nested function. 

## Data Range
```{r}
dataRanges <- sapply(1:length(penguins_data), function(x) ifelse(
  class(penguins_data[[x]]) == "factor",
  paste0(levels(penguins_data[[x]]), collapse = ", "),
  paste0(range(penguins_data[[x]], na.rm = TRUE), collapse = ", ")
))

view(dataRanges)
```

Using double square brackets [[]] extracts the actual value.The levels and range functions return a vector of multiple values (min and max), so we 

Paste or paste0 can be used. Paste, pastes (concatenates) multiple things together. Paste inlcudes a space between concatenated objects, paste0 does not include a space between objects. Here we added the 'collapse' agrument to insert a ", " between object. the list() function works similarly. 

## Combine into Data dictionary 

```{r}
penguins_dd <- data.frame(
  varNames,
  dataClass,
  dataType,
  observationsWithValues,
  dataRanges
)

view(penguins_dd)
```

The functions we created for each variable could be included within the penguins_dd creation. That way it can all be run at once. See below. 

```{r}
penguins_dd <-
data.frame(
  varNames,
  dataClass,
  dataType,
  observationsWithValues,
  dataRanges
)
```

# Generalization 

We want our code to be usable by others. So we can make it into a function so other data sets can be used. We will also include some error checking to make sure the correct input is given. 

```{r}
datadictionary_fn <- function(dataset){
  # test if its a data frame and stop if it's not 
  if(!"data.frame" %in% class(dataset)){
    stop("This is not a data frame.")
  }
  return(
    data.frame(
  varNames = names(dataset),
  dataClass = unname(sapply(dataset, class)),
  dataType = unname(sapply(dataset, typeof)),
  observationsWithValues = unname(sapply(dataset, function(x) sum(!is.na(x)))),
  dataRanges = sapply(1:length(dataset), function(x) ifelse(
  class(dataset[[x]]) == "factor",
  paste0(levels(dataset[[x]]), collapse = ", "),
  paste0(range(dataset[[x]], na.rm = TRUE), collapse = ", ")
))
)
  )
}

penguins_dd <- datadictionary_fn(penguins_data)
```

Test with a non-data frame dataset 

```{r}
test_var <- c(1:5)
test_dd <- datadictionary_fn(test_var)
```

