---
title: "Filter And Select"
author: "Naomi Loughiln"
date: "2025-05-14"
output: html_document
---
# Filter and Select

## Setup and Load the Dataset 

```{r setup, include=FALSE}
library(dplyr)
js_data_path <- "data/timeuse_day3_1.Rdata"

load(js_data_path)
```

## Filtering 

Filter for distinct/unique values in the feelRushed column to see what categories exist. 

```{r}
js_data |> 
  distinct(feelRushed)
``` 

We can also look at distributions using the count() and arrange() functions.

```{r}
# Sleep Duration distribution 
js_data |> 
  count(durSleep) |> 
  arrange(desc(n)) |> 
  head(10) 
```


```{r}
# Work Duration Distriburion
js_data |> 
  count(durWork) |> 
  arrange(desc(n)) |> 
  head(10)
```

## Basic Filtering 

Filter the feelRushed column for those who responded with feelin rushed daily (which corresponds to a value of 1). This shows all of the subjects who feel rushed daily. 

```{r}
js_data |> 
  filter(feelRushed == 1) |> 
  head()
```


We can also use logical operators to filter for certain values. 

```{r}
# Filter for sleep duration less than 600
js_data |> 
  filter(durSleep < 600) |>
  head()
```

```{r}
# Filter for sleep duration between 600 and 1000
js_data |> 
  filter(durSleep >= 600 & durSleep <= 1000) |>
  head()
```


## Complex Filtering 


Using range comparison (greater or equal). 

```{r}
js_data |> 
  filter(feelRushed <= 3) |>
  head()
```

We can use the "&" to specify multiple conditions. 

```{r}
js_data |> 
  filter(feelRushed > 3 & feelRushed <= 6) |>
  head()
```

Calculate how many rows remain after filtering 

`js_data |> 
  filter(feelRushed > 3 & feelRushed <= 6) |>
  head()
  
```{r}
all_rows <- js_data |> 
  nrow()
rushed_rows <- js_data |> 
  filter(feelRushed <= 3) |>
  nrow()
not_rushed_rows <- js_data |> 
  filter(feelRushed > 3 & feelRushed <= 6) |>
  nrow()

print(paste("The number of rows in data is:", all_rows))

print(paste("The number of rows in rushed is:", rushed_rows))

print(paste("The number of rows in not rushed is:", not_rushed_rows))
```

Chain Multiple Conditions using the pipe operator 

```{r}
js_data |> 
  filter(feelRushed == 1 | feelRushed == 2 | feelRushed == 3) |> 
  head()
```


Calculate how many rows remain after filtering 

```{r}
rushed_rows <- js_data |> 
  filter(feelRushed == 1 | feelRushed == 2 | feelRushed == 3) |>
  nrow()
not_rushed_rows <- js_data |> 
  filter(feelRushed == 4 | feelRushed == 5 | feelRushed == 6) |>
  nrow()

print(paste("The number of rows in rushed is:", rushed_rows))

print(paste("The number of rows in not rushed is:", not_rushed_rows))
```

## Select Relevant Variables 

We’ll select only these three columns from our rushed and not_rushed datasets using select().

```{r}
rushed_time <- js_data |> 
  filter(feelRushed <= 3) |>
  select(durWork, durSleep, durAlone) 

not_rushed_time <- js_data |> 
  filter(feelRushed > 3) |> 
  select(durWork, durSleep, durAlone)


rushed_time |> 
  head()

not_rushed_time |> 
  head()
```


## Compare Time Use 


```{r}
mean_rushed <- rushed_time |> 
  summarise(
    durSleep = mean(durSleep, na.rm = TRUE),
    durAlone = mean(durAlone, na.rm = TRUE),
    durWork = mean(durWork, na.rm = TRUE)
  )

mean_not_rushed <- not_rushed_time |> 
  summarise(
    durSleep = mean(durSleep, na.rm = TRUE),
    durAlone = mean(durAlone, na.rm = TRUE),
    durWork = mean(durWork, na.rm = TRUE)
  )
```

We can do the same thing mmore efficiently, in less lines of code: 

```{r}
# Calculate the mean values for each variable using summarise and across
mean_rushed <- rushed_time |> 
  summarise(across(everything(), ~ mean(. , na.rm = TRUE)))

mean_not_rushed <- not_rushed_time |> 
  summarise(across(everything(), ~ mean(. , na.rm = TRUE)))
```

We calculate the difference in means between these two groups.The subtraction (mean_rushed - mean_not_rushed) gives a new one‐row tibble showing how much more (or less) time “rushed” individuals spend on each activity compared to “not rushed” individuals.

```{r}
diff_means <- mean_rushed - mean_not_rushed
```

## Recode and Save

```{r}
js_data |>
  count(popCenter)
```

We can use dplyr’s mutate() along with if_else() to turn those numbers into descriptive labels:

```{r}

js_data <- js_data |> mutate(
  popCenter=if_else(popCenter==1,'urban',
                    if_else(popCenter==2,'rural','PEI'))
)
```

Next, we’ll create a new flag called isFeelRushed to mark whether each respondent feels rushed or not. This will be useful for our upcoming data-visualization tutorial:

```{r}
js_data <- js_data |> mutate(
  isFeelRushed=if_else(feelRushed <= 3,1,0)
)

```

### SaVE

```{r}
save(js_data, file = "../data/time_use_day3_2.RData")

```




